---
title: "22q_subcort_volumes"
author: "charlie schleifer"
date: "2/8/23"
output: html_document
---
## Overview
Subcortical nuclei volumes for Thalamus, Hippocampus, and Amygdala were generated with FreeSurfer 7.3.2 segment_structures
This script is for analysis of those results in 22qDel, 22qDup, and HCS

## set up workspace
```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# clear workspace
rm(list = ls(all.names = TRUE))

# list packages to load
# ciftiTools dependency rgl may need XQuartz installed in order to visualize surfaces
#packages <- c("conflicted", "here", "magrittr", "mgcv", "gratia", "lme4", "lmerTest", "invgamma", "longCombat", "ciftiTools", "readxl", "dplyr", "data.table", "DescTools","tableone", "tibble", "reshape2", "viridis", "scico", "ggplot2", "gridExtra", "ggpubr","stringr")

packages <- c("conflicted", "here", "magrittr", "dplyr", "data.table", "readxl", "tableone","ggplot2", "viridis", "longCombat")

# install packages if not yet installed
all_packages <- rownames(installed.packages())
installed_packages <- packages %in% all_packages
if (any(installed_packages == FALSE)){install.packages(packages[!installed_packages])}

# load packages
invisible(lapply(packages, library, character.only = TRUE))

# use the filter function from dplyr, not stats
conflict_prefer("filter", "dplyr")

# get path to project repo directory
project <- here()
print(paste("Project directory:", project))

```

## read data
```{r}
# path to data (organized with subject directories containing only the relevant files)
# TODO: put data in a shared location
dpath <- "~/Dropbox/PhD/bearden_lab/22q/analyses/subcort_smri/"

# get files
sessions_all <- list.files(path=dpath, pattern="Q_[0-9]")

# function to read text file and change rows to columns
read_fssubcort <- function(path, sesh, name){
  file <- file.path(dpath,sesh,name)
  orig <- read.table(file)
  out <- as.data.frame(t(orig$V2))
  names(out) <- orig$V1
  #out$MRI_S_ID <- sesh
  # for longitudinal 
  out$MRI_S_ID <- gsub(".long.*","", sesh)
  return(out)
}

read_fssubcort_long <- function(path, sesh, name){
  file <- file.path(dpath,sesh,name)
  orig <- read.table(file)
  out <- as.data.frame(t(orig$V2))
  names(out) <- orig$V1
  #out$MRI_S_ID <- sesh
  # for longitudinal need to remove end of folder name to get MRI_S_ID
  out$MRI_S_ID <- gsub(".long.*","", sesh)
  return(out)
}

# read data
thal_lr <- lapply(sessions_all, function(s) read_fssubcort_long(path=dpath,sesh=s,name="ThalamicNuclei.long.volumes.txt")) %>% do.call(rbind,.)
names(thal_lr)[which(!names(thal_lr)=="MRI_S_ID")] <- paste0("Thal_",names(thal_lr)[which(!names(thal_lr)=="MRI_S_ID")])
amy_l <- lapply(sessions_all, function(s) read_fssubcort_long(path=dpath,sesh=s,name="lh.amygNucVolumes.long.txt")) %>% do.call(rbind,.)
names(amy_l)[which(!names(amy_l)=="MRI_S_ID")] <- paste0("Amy_Left_",names(amy_l)[which(!names(amy_l)=="MRI_S_ID")])
amy_r <- lapply(sessions_all, function(s) read_fssubcort_long(path=dpath,sesh=s,name="rh.amygNucVolumes.long.txt")) %>% do.call(rbind,.)
names(amy_r)[which(!names(amy_r)=="MRI_S_ID")] <- paste0("Amy_Right_",names(amy_r)[which(!names(amy_r)=="MRI_S_ID")])
hip_l <- lapply(sessions_all, function(s) read_fssubcort_long(path=dpath,sesh=s,name="lh.hippoSfVolumes.long.txt")) %>% do.call(rbind,.)
names(hip_l)[which(!names(hip_l)=="MRI_S_ID")] <- paste0("Hip_Left_",names(hip_l)[which(!names(hip_l)=="MRI_S_ID")])
hip_r <- lapply(sessions_all, function(s) read_fssubcort_long(path=dpath,sesh=s,name="rh.hippoSfVolumes.long.txt")) %>% do.call(rbind,.)
names(hip_r)[which(!names(hip_r)=="MRI_S_ID")] <- paste0("Hip_Right_",names(hip_r)[which(!names(hip_r)=="MRI_S_ID")])

# merge data
subcort_all <- merge(x=thal_lr, y=amy_l, by="MRI_S_ID")
subcort_all <- merge(x=subcort_all, y=amy_r, by="MRI_S_ID")
subcort_all <- merge(x=subcort_all, y=hip_l, by="MRI_S_ID")
subcort_all <- merge(x=subcort_all, y=hip_r, by="MRI_S_ID")

# replate problematic characters with underscores in column names
colnames(subcort_all) <- gsub("-","_",colnames(subcort_all))
colnames(subcort_all) <- gsub("\\(","_",colnames(subcort_all))
colnames(subcort_all) <- gsub("\\)","_",colnames(subcort_all))

# read eTIV
etiv <- read.csv(file.path(dpath,"etiv_cross_sectional.txt"), header=TRUE)

```

## load sistat data and get lists of scans to use
all sistat tables should be exported as CSVs into a single directory
the next several chunks deal with reading, cleaning and annotating the data exported from sistat, and then age matching
the hcs sample is younger than del due to a large amount of very young hcs subjects. plan is to match samples by using followup timepoints rather than baseline for some younger participants, and dropping several older del subjects, and younger hcs subjects (prioritizing dropping subjects with worse motion stats when possible)
```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# set location of directory with ucla sistat CSVs
csvdir_ucla <- file.path(project,"demographics/ucla_sistat")

# get list of files_ucla in directory
files_ucla <- list.files(csvdir_ucla)
fpaths <- lapply(files_ucla, function(file) paste(csvdir_ucla,file,sep="/"))

# clean names
fnames <- gsub(".csv","",files_ucla)
fnames <- gsub("Re22Q_","",fnames)
fnames <- gsub("Form_","",fnames)
fnames <- gsub("Qry_","",fnames)

# read all, set to na: "-9999", "-9998","." 
input_all_ucla <- lapply(fpaths, read.csv, header=T, na.strings=c(".","-9999","-9998"), strip.white=T, sep=",")
names(input_all_ucla) <- fnames
df_all_ucla <- lapply(input_all_ucla, function(x) data.frame(x))

# get demo_mri 
ucla_demo <- df_all_ucla$demo_mri

# remove "FAMILY MEMBER" designation from subject identity
ucla_demo$SUBJECT_IDENTITY <- ucla_demo$SUBJECT_IDENTITY %>% sub("FAMILY MEMBER","",.) %>% sub(",","",.) %>% trimws(which="both") %>% as.factor
# change sex coding from 0/1 to F/M and set to factor
ucla_demo$SEX <- factor(ucla_demo$SEX,levels=c(0,1),labels=c("F","M"))
```

# TEMPORARY
read temporary csv with several subjects not yet in sistat
```{r}
# TODO: this chunk is temporary until sistat is updated 
# TODO: note: q_0526 sex was "na" in original sheet, changed to M because combat can't have NAs
# read new data
temp_demo <- read_xlsx(file.path(project,"demographics/temporary/sMRI_demo_info_forCharlie.xlsx"), col_names=TRUE,na="",trim_ws = TRUE)

# make empty demographics data frame to add new data to
demo_add <- ucla_demo[1:nrow(temp_demo),]
demo_add[,] <- NA
demo_add$SUBJECTID <- temp_demo$`Subject ID`
demo_add$SUBJECT_IDENTITY <- temp_demo$Diagnosis
demo_add$MRI_S_ID <- temp_demo$`MRI ID`
demo_add$SEX <- as.factor(temp_demo$Sex)
demo_add$AGE <- temp_demo$Age
demo_add$AGEMONTH <- temp_demo$Age*12
demo_add$CONVERTEDVISITNUM <- 2

# append to ucla demo
ucla_demo <- rbind(ucla_demo,demo_add)
```
 
continue regular steps
```{r}
# manually fix missing sex for Q_0381_09102019
# TODO: fix in sistat and re-export
ucla_demo[which(ucla_demo$MRI_S_ID == "Q_0381_09102019"),"SEX"] <- "F"

# set race=NA to 7 (unknown)
ucla_demo$RACE[is.na(ucla_demo$RACE)] <- 7
# set race as factor 1=American Indian/Alaska Native; 2=Asian; 3=Native Hawaiian/Pacific Islander; 4=Black or African American; 5=White; 6=Multiple; 7=Unknown
ucla_demo$RACE <- factor(ucla_demo$RACE,levels=c(1:7),labels=c("1_Native_American","2_Asian","3_Pacific_Island","4_Black","5_White","6_Multiple","7_Unknown"))
# ethnicity as factor with 0=N 1=Y
ucla_demo$HISPANIC[is.na(ucla_demo$HISPANIC)] <- "Unknown"
ucla_demo$HISPANIC <- factor(ucla_demo$HISPANIC,levels=c(0,1,"Unknown"),labels=c("N","Y","Unknown"))
# get more accurate age with AGEMONTH/12
ucla_demo$AGE <- as.numeric(ucla_demo$AGEMONTH)/12 

# subset to used scans
ucla_demo <- filter(ucla_demo, MRI_S_ID %in% subcort_all$MRI_S_ID)

# function to add column to code timepoints relative to sample used (i.e. if visit 1 and 1.12 missing, then 1.24 is baseline)
# trio/prisma coded as T/P-visit_n where T-1 would be the subject's first trio scan and P-1 the first prisma, P-2 the second...
# function should be applied to the indicies of rows (r) in a subset of demo_mri
gettp <- function(r, df){
  sub <- df$SUBJECTID[[r]]
  visit <- df$CONVERTEDVISITNUM[[r]]
  all_visits <- df$CONVERTEDVISITNUM[which(df$SUBJECTID == sub)] %>% sort
  n_visits <- length(all_visits)
  nt_visits <-length(which(all_visits < 2))
  np_visits <- length(which(all_visits >= 2))
  visit_index <- which(all_visits == visit)
  if (visit < 2){
    label=paste("T-",visit_index,sep="")
  }else if (visit >= 2){
    p_visits <- all_visits[which(all_visits >= 2)] %>% sort
    p_visit_index <- which(p_visits == visit)
    label=paste("P-",p_visit_index,sep="")
  }
  return(c(sub,visit,label,n_visits,nt_visits,np_visits,visit_index))
}

# get timepoints
timepoints <- lapply(1:nrow(ucla_demo),function(r) gettp(r,ucla_demo)) %>% do.call(rbind,.) %>% as.data.frame
colnames(timepoints) <- c("SUBJECTID","CONVERTEDVISITNUM","converted_timepoint","n_timepoints","n_trio","n_prisma","visit_index")
ucla_demo_tp <- cbind(ucla_demo,timepoints[,3:7])
ucla_demo_tp$visit_index %<>% as.factor

# subset to BASELINE under max age limit (45 years old)
ucla_demo_tp_agelim <- filter(ucla_demo_tp, ucla_demo_tp$AGE < 50 & visit_index ==1)

# subset to hcs del
#ucla_demo_hcs_del <- ucla_demo_tp_agelim %>% filter(SUBJECT_IDENTITY=="CONTROL" | SUBJECT_IDENTITY =="PATIENT-DEL")

# remove unused factor levels
ucla_demo_tp_agelim %<>% droplevels
```

All timepoints, pre-matching demographics summary
```{r}
demo_summary <- CreateTableOne(data=ucla_demo_tp_agelim,vars=c("AGE","SEX"),strata="SUBJECT_IDENTITY",addOverall=F)
print(demo_summary, showAllLevels=T)
```

Baseline pre-matching summary
```{r}
demo_summary_bl <- CreateTableOne(data=filter(ucla_demo_tp_agelim, visit_index == 1),vars=c("AGE","SEX"),strata="SUBJECT_IDENTITY",addOverall=F)
print(demo_summary_bl)
```
## Harmonize sites
longCombat for subcortical structures 
```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# add site column
#ucla_demo_tp_agelim$site <- gsub("*-[0-9]","",ucla_demo_tp_agelim$converted_timepoint)
#
## get feature names
#sc_names <- names(subcort_all)[which(names(subcort_all) != "MRI_S_ID")]
#
## merge subcort_all with demo
#demo_sc <- merge(x=ucla_demo_tp_agelim, y=subcort_all, by="MRI_S_ID", all.x=TRUE)
#
## merge with eTIV
#demo_sc <- merge(x=demo_sc, y=etiv, by="MRI_S_ID")
#
## make numeric gene dosage column from SUBJECT_IDENTITY
#demo_sc$gene_dosage <- demo_sc$SUBJECT_IDENTITY %>% gsub("PATIENT-DEL","1",.) %>% gsub("CONTROL","2",.) %>% gsub("PATIENT-DUP","3",.) %>% as.numeric
#
## set up longCombat variables
## formula should match fixed effects in your subsequent analysis
## subject id coded as random effect by (1|subject id variable)
#demovars <- c("MRI_S_ID","SUBJECTID","site","gene_dosage","AGE","SEX","eTIV","visit_index")
#features <- sc_names
#idvar <- 'MRI_S_ID'
#batchvar <- 'site'
#timevar <- 'visit_index'
#formula <- 'gene_dosage + AGE + SEX + eTIV'
#ranef <- '(1|SUBJECTID)'
#
## make data frame with columns for each variable in the model
## one column for each variable in your formula as well as one column for each neuroimaging feature
## input df should not have any unused columns or package will error
## one row per unique scan
#combat_input<- demo_sc[,c(demovars,features)]
#
## run longCombat
#sc_vol_combat <- longCombat(data=combat_input, idvar=idvar, timevar=timevar, batchvar=batchvar, features=features, formula=formula, ranef=ranef)
#
## get the harmonized data
#sc_vol_combat_data <- sc_vol_combat$data_combat
#
## merge combat back with original
#demo_combat <- merge(x=demo_sc, y=sc_vol_combat_data, by=c("MRI_S_ID","visit_index","site"))
```

LME to test effect of 22q11.2 CNV dosage on combat adjusted features
```{r warning=FALSE}
# add site column
ucla_demo_tp_agelim$site <- gsub("*-[0-9]","",ucla_demo_tp_agelim$converted_timepoint)

# get feature names
sc_names <- names(subcort_all)[which(names(subcort_all) != "MRI_S_ID")]

# merge subcort_all with demo
demo_sc <- merge(x=ucla_demo_tp_agelim, y=subcort_all, by="MRI_S_ID", all.x=TRUE)

# merge with eTIV
demo_sc <- merge(x=demo_sc, y=etiv, by="MRI_S_ID")

# make numeric gene dosage column from SUBJECT_IDENTITY
demo_sc$gene_dosage <- demo_sc$SUBJECT_IDENTITY %>% gsub("PATIENT-DEL","1",.) %>% gsub("CONTROL","2",.) %>% gsub("PATIENT-DUP","3",.) %>% as.numeric


# test random intercept model at each region
lm_nocombat <- lapply(sc_names, function(r) lm(formula=reformulate(c("gene_dosage","AGE","SEX","eTIV","site"), response=r),data=demo_sc))

# get p-val for gene dosage effect
gene_dosage_effect <- lapply(lm_nocombat, function(l) summary(l)$coefficients["gene_dosage",c("Estimate","Pr(>|t|)")]) %>% do.call(rbind,.) %>% as.data.frame 
colnames(gene_dosage_effect) <- c("gene_dosage_beta","gene_dosage_p")
gene_dosage_effect$gene_dosage_fdr_q <- p.adjust(gene_dosage_effect$gene_dosage_p, method="fdr")
gene_dosage_effect$Region <- sc_names
gene_dosage_effect$fdr_sig <- gene_dosage_effect$gene_dosage_fdr_q < 0.05

# get only significant effects
gene_dosage_effect_sig <- filter(gene_dosage_effect,fdr_sig==TRUE)
gene_dosage_effect_sig <- gene_dosage_effect_sig[order(gene_dosage_effect_sig$gene_dosage_beta),]

#write.table(gene_dosage_effect_sig, file="~/Dropbox/PhD/bearden_lab/22q/analyses/subcort_smri/gene_dosage_lme_significant.csv", sep=",", col.names = TRUE, row.names = FALSE)
```

volume barplots by gene dosage for select regions
```{r}
ggplot(data=demo_combat, aes(x=gene_dosage, y=Thal_Left_MDm.combat, group=gene_dosage, fill=gene_dosage))+
  geom_boxplot()+
  scale_fill_viridis()+
  xlab("22q11.2 CNV dosage")+
  ylab("L Thal MDm volume")+
  ggtitle("Left Thal MDm volume")+
  theme_classic()


ggplot(data=demo_combat, aes(x=gene_dosage, y=Thal_Right_MDm.combat, group=gene_dosage, fill=gene_dosage))+
  geom_boxplot()+
  scale_fill_viridis()+
  xlab("22q11.2 CNV dosage")+
  ylab("R Thal MDm volume")+
  ggtitle("Right Thal MDm volume")+
  theme_classic()

ggplot(data=demo_combat, aes(x=gene_dosage, y=Hip_Right_Whole_hippocampus.combat, group=gene_dosage, fill=gene_dosage))+
  geom_boxplot()+
  scale_fill_viridis()+
  xlab("22q11.2 CNV dosage")+
  ylab("R Whole Hipp volume")+
  ggtitle("Right whole Hipp volume")+
  theme_classic()


```
To test for group differences, not just gene dosage effects (e.g. Dup and Del both decreased), use a model with a binary variable for each group? or two separate case/control models? 

```{r warning=FALSE}
# test random intercept model at each region separately in Del and Dup
lme_combat_del <- lapply(sc_names_combat, function(r) lmerTest::lmer(formula=reformulate(c("gene_dosage","AGE","SEX","eTIV","site","(1|SUBJECTID)"), response=r),data=filter(demo_combat, gene_dosage %in% c(1,2)), REML=TRUE))

lme_combat_dup <- lapply(sc_names_combat, function(r) lmerTest::lmer(formula=reformulate(c("gene_dosage","AGE","SEX","eTIV","site","(1|SUBJECTID)"), response=r),data=filter(demo_combat, gene_dosage %in% c(2,3)), REML=TRUE))

# get p-val for gene dosage effect
gene_dosage_effect_del <- lapply(lme_combat_del, function(l) summary(l)$coefficients["gene_dosage",c("Estimate","Pr(>|t|)")]) %>% do.call(rbind,.) %>% as.data.frame 
colnames(gene_dosage_effect_del) <- c("gene_dosage_beta","gene_dosage_p")
gene_dosage_effect_del$Region <- sc_names_combat
gene_dosage_effect_del$model <- "Del"

gene_dosage_effect_dup <- lapply(lme_combat_dup, function(l) summary(l)$coefficients["gene_dosage",c("Estimate","Pr(>|t|)")]) %>% do.call(rbind,.) %>% as.data.frame 
colnames(gene_dosage_effect_dup) <- c("gene_dosage_beta","gene_dosage_p")
gene_dosage_effect_dup$Region <- sc_names_combat
gene_dosage_effect_dup$model <- "Dup"

del_dup_vs_hc <- rbind(gene_dosage_effect_del,gene_dosage_effect_dup)
del_dup_vs_hc$gene_dosage_fdr_q <- p.adjust(del_dup_vs_hc$gene_dosage_p, method="fdr")
del_dup_vs_hc$fdr_sig <- del_dup_vs_hc$gene_dosage_fdr_q < 0.05

del_dup_vs_hc_merged <- merge(x=filter(del_dup_vs_hc, model=="Del"), y=filter(del_dup_vs_hc, model=="Dup"), by="Region")

# get only effects that are significant in both models
del_dup_vs_hc_merged_sig <- filter(del_dup_vs_hc_merged,fdr_sig.x==TRUE,fdr_sig.y==TRUE)
#del_dup_vs_hc_sig <- del_dup_vs_hc_sig[order(del_dup_vs_hc_sig$gene_dosage_beta),]
```

LME for gene_dosage * Age linear interaction
```{r warning=FALSE}
# test random intercept model for group*age interaction at each region
lme_combat_agegene <- lapply(sc_names_combat, function(r) lmerTest::lmer(formula=reformulate(c("gene_dosage*AGE","SEX","eTIV","site","(1|SUBJECTID)"), response=r),data=demo_combat, REML=TRUE))

# get p-val for gene dosage effect
gene_by_age_effect <- lapply(lme_combat_agegene, function(l) summary(l)$coefficients["gene_dosage:AGE",c("Estimate","Pr(>|t|)")]) %>% do.call(rbind,.) %>% as.data.frame 
colnames(gene_by_age_effect) <- c("gene_by_age_beta","gene_by_age_p")
gene_by_age_effect$gene_by_age_fdr_q <- p.adjust(gene_by_age_effect$gene_by_age_p, method="fdr")
gene_by_age_effect$Region <- sc_names_combat
gene_by_age_effect$fdr_sig <- gene_by_age_effect$gene_by_age_fdr_q < 0.05

# get only significant effects
gene_by_age_effect_sig <- filter(gene_by_age_effect,fdr_sig==TRUE)
#gene_by_age_effect_sig <- gene_by_age_effect_sig[order(gene_by_age_effect_sig$gene_dosage_beta),]

#write.table(gene_dosage_effect_sig, file="~/Dropbox/PhD/bearden_lab/22q/analyses/subcort_smri/gene_dosage_lme_significant.csv", sep=",", col.names = TRUE, row.names = FALSE)
```

LME for separate group*age interaction in each group 
```{r warning=FALSE}

# test random intercept model for group*age interaction at each region
lme_combat_agegene_del <- lapply(sc_names_combat, function(r) lmerTest::lmer(formula=reformulate(c("gene_dosage*AGE","SEX","eTIV","site","(1|SUBJECTID)"), response=r),data=filter(demo_combat, gene_dosage %in% c(1,2)), REML=TRUE))

lme_combat_agegene_dup <- lapply(sc_names_combat, function(r) lmerTest::lmer(formula=reformulate(c("gene_dosage*AGE","SEX","eTIV","site","(1|SUBJECTID)"), response=r),data=filter(demo_combat, gene_dosage %in% c(2,3)), REML=TRUE))

# get p-val for gene dosage effect
gene_by_age_effect_del <- lapply(lme_combat_agegene_del, function(l) summary(l)$coefficients["gene_dosage:AGE",c("Estimate","Pr(>|t|)")]) %>% do.call(rbind,.) %>% as.data.frame 
colnames(gene_by_age_effect_del) <- c("gene_by_age_beta","gene_by_age_p")
gene_by_age_effect_del$Region <- sc_names_combat
gene_by_age_effect_del$model <- "Del"

gene_by_age_effect_dup <- lapply(lme_combat_agegene_dup, function(l) summary(l)$coefficients["gene_dosage:AGE",c("Estimate","Pr(>|t|)")]) %>% do.call(rbind,.) %>% as.data.frame 
colnames(gene_by_age_effect_dup) <- c("gene_by_age_beta","gene_by_age_p")
gene_by_age_effect_dup$Region <- sc_names_combat
gene_by_age_effect_dup$model <- "Dup"

gene_by_age_del_dup_vs_hc <- rbind(gene_by_age_effect_del,gene_by_age_effect_dup)
gene_by_age_del_dup_vs_hc$gene_by_age_fdr_q <- p.adjust(gene_by_age_del_dup_vs_hc$gene_by_age_p, method="fdr")
#gene_by_age_del_dup_vs_hc$fdr_sig <- gene_by_age_del_dup_vs_hc$gene_dosage_fdr_q < 0.05

#gene_by_age_del_dup_vs_hc_merged <- merge(x=filter(gene_by_age_del_dup_vs_hc, model=="Del"), y=filter(gene_by_age_del_dup_vs_hc, model=="Dup"), by="Region")

# get only effects that are significant in both models
#del_dup_vs_hc_merged_sig <- filter(del_dup_vs_hc_merged,fdr_sig.x==TRUE,fdr_sig.y==TRUE)



```







